---
title: "treeR_ch12"
author: "Ricarda Scholte-Reh"
date: "2024-10-12"
homepage: "https://Ricarda-Scholte-Reh.github.io"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
---

# Generating temperature scenarios

## Exercises
1.  For the location you chose for your earlier analyses, use chillR's weather generator to produce 100 years of synthetic temperature data.
```{r}
#rm(list=ls()) # clears the environment from unwanted values etc.
library(chillR)
library(tidyverse)
library(ggplot2)
library(tibble)
library(magrittr)

patched_complete <- read.csv("D:/! Ricardas Daten/Studium/Master/03. Semester (WiSe 24-25)/Tree phenology analysis in R/Logbook chapters/patched_weather_3_Wesel.csv")

# we simulate here with the data form 1998 to 2009 100 times how the weather could have been in this time
Temp <- patched_complete %>%
  temperature_generation(years = c(1998,2009),
                         sim_years = c(2001,2100))


Temperatures <- patched_complete %>% filter(Year %in% 1998:2009) %>%
  cbind(Data_source = "observed") # cbind() adds more coloumns to the table

new_data <- Temp[[1]] %>%
  select(Year, Month, Day, Tmin, Tmax) %>%  # Nur die gewünschten Spalten auswählen
  mutate(Data_source = "simulated")        # Neue Spalte Data_source hinzufügen

# Daten an Temperatures anhängen
Temperatures <- bind_rows(Temperatures, new_data)


Temperatures <- Temperatures %>% 
  mutate(Date = as.Date(ISOdate(2000, Month, Day))) # mutate creates a coloumn from already existing data
```

```{r}
ggplot(data = Temperatures,
       aes(Date,
           Tmin)) +
  geom_smooth(aes(colour = factor(Year))) +
  facet_wrap(vars(Data_source)) +
  theme_bw(base_size = 20) +
  theme(legend.position = "none") +
  scale_x_date(date_labels = "%b")

ggplot(data = Temperatures,
       aes(Date,
           Tmax)) + # creates a empty coordinate system
  geom_smooth(aes(colour = factor(Year))) + # creates a smoothing regression to make the data better obsevable (alternative is geom_line() then you getalso the nice colors but without the smoothing evect)
  facet_wrap(vars(Data_source)) + # 
  theme_bw(base_size = 20) +
  theme(legend.position = "none") + # deletes the legend (is here only confusing)
  scale_x_date(date_labels = "%b") # shows on the xachsis only the month without the year

```


2. Calculate winter chill (in Chill Portions) for your synthetic weather, and illustrate your results as histograms and cumulative distributions.
```{r}
chill_observed <- Temperatures %>%
  filter(Data_source == "observed") %>%
  stack_hourly_temps(latitude = 50.4) %>%
  chilling(Start_JDay = 305,
           End_JDay = 59)
  
chill_simulated <- Temperatures %>%
  filter(Data_source == "simulated") %>%
  stack_hourly_temps(latitude = 50.4) %>%
  chilling(Start_JDay = 305,
           End_JDay = 59)
  
chill_comparison <-
  cbind(chill_observed,
        Data_source = "observed") %>%
  rbind(cbind(chill_simulated,
              Data_source = "simulated"))
  
chill_comparison_full_seasons <- 
  chill_comparison %>%
  filter(Perc_complete == 100)

chill_comparison_full_seasons$Data_source <- factor(
  chill_comparison_full_seasons$Data_source,
  levels = c("simulated", "observed")
)


chill_comparison_full_seasons$Data_source <- factor(
  chill_comparison_full_seasons$Data_source,
  levels = c("simulated", "observed")
)

chill_comparison_full_seasons <- chill_comparison_full_seasons[order(
  chill_comparison_full_seasons$Data_source), ]
```

```{r}
ggplot(chill_comparison_full_seasons,
       aes(x = Chill_portions)) + 
  geom_histogram(binwidth = 1, position = "identity",
                 aes(fill = factor(Data_source))) +
  scale_fill_brewer(palette="Set2") +
  theme_bw(base_size = 20) +
  labs(fill = "Data source") +
  xlab("Chill accumulation (Chill Portions)") +
  ylab("Frequency")
```

```{r}
chill_simulations <-
  chill_comparison_full_seasons %>%
  filter(Data_source == "simulated")
  
ggplot(chill_simulations,
       aes(x = Chill_portions)) +
  stat_ecdf(geom = "step",
            lwd = 1.5,
            col = "blue") +
  ylab("Cumulative probability") +
  xlab("Chill accumulation (in Chill Portions)") +
  theme_bw(base_size = 20)
```

3. Produce similar plots for the number of freezing hours (<0°C) in April (or October, if your site is in the Southern Hemisphere) for your location of interest.
```{r}
# filter only our observed data and calculate with the coordinates (latitude) the chilling day 
chill_observed <- Temperatures %>%
  filter(Data_source == "observed") %>%
  stack_hourly_temps(latitude = 51.4) %>%
  chilling(Start_JDay = 305,
           End_JDay = 59)


### adding a freezing model - to make this for April, you'll also have to adjust the dates in the calculations
df <- data.frame(
  lower= c(-1000, 0),
  upper= c(    0, 1000),
  weight=c(    1, 0))

freezing_hours <- function(x) step_model(x,df)

freezing_hours(c(1, 2, 4, 5, -10))

chill_observed <- Temperatures %>%
  filter(Data_source == "observed") %>%
  stack_hourly_temps(latitude = 51.4) %>%
  tempResponse(Start_JDay = 305, # more flexible than the chilling() function, we feed it our freezing function
               End_JDay = 59,
               models = list(Frost = freezing_hours,
                             Chill_portions = Dynamic_Model,
                             GDH = GDH))


####

# repeating the freezing function with the simulated data
chill_simulated <- Temperatures %>%
  filter(Data_source == "simulated") %>%
  stack_hourly_temps(latitude = 51.4) %>%
  tempResponse(Start_JDay = 305,
               End_JDay = 59,
               models=list(Frost = freezing_hours,
                           Chill_portions = Dynamic_Model,
                           GDH = GDH))

# combine the freezing data for the observed and the simulated
chill_comparison <-
  cbind(chill_observed,
        Data_source = "observed") %>%
  rbind(cbind(chill_simulated,
              Data_source = "simulated"))

# sometimes there is data missing for the winter and the influence the data negative, so we remove them
chill_comparison_full_seasons <-
  chill_comparison %>%
  filter(Perc_complete == 100)


chill_comparison_full_seasons$Data_source <- factor(
  chill_comparison_full_seasons$Data_source,
  levels = c("simulated", "observed")
)

chill_comparison_full_seasons <- chill_comparison_full_seasons[order(
  chill_comparison_full_seasons$Data_source), ]
```


```{r}
# plot the data as a histogram
ggplot(chill_comparison_full_seasons,
       aes(x = Frost)) + 
  geom_histogram(binwidth = 25, position = "identity",
                 aes(fill = factor(Data_source))) +
  scale_fill_brewer(palette="Set2") +
  theme_bw(base_size = 10) +
  labs(fill = "Data source") +
  xlab("Frost incidence during winter (hours)") +
  ylab("Frequency")

chill_simulations <-
  chill_comparison_full_seasons %>%
  filter(Data_source == "simulated")

# shows how many frost hours are how likely to happen
ggplot(chill_simulations,
       aes(x = Frost)) +
  stat_ecdf(geom = "step", # accumulated density function
            lwd = 1.5,
            col = "blue") +
  ylab("Cumulative probability") +
  xlab("Frost incidence during winter (hours)") +
  theme_bw(base_size = 20)

# Here's the amount of chill that is exceeded in 90% of all years.
quantile(chill_simulations$Chill_portions, 0.1)

# and here's the 50% confidence interval (25th to 75th percentile)
quantile(chill_simulations$Chill_portions, c(0.25, 0.75))

```



