---
title: "treeR_ch10"
author: "Ricarda Scholte-Reh"
date: "2024-10-12"
homepage: "https://Ricarda-Scholte-Reh.github.io"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
---

# Getting temperature data

## Exercises

1. Choose a location of interest and find the 25 closest weather stations using the handle_gsod function.

- location of interest: Dairy farm with own fodder cultivation (belongs to my uncle)
                        Bislich, 46487 Wesel
- coordinates: latitude: 51°40'58.4"N
               longitude: 6°29'46.4"E
```{r}
rm(list=ls()) # clears the environment from unwanted values etc.
library(chillR)
library(tidyverse)
library(ggplot2)
library(tibble)
library(magrittr)

station_list <- handle_gsod(action = "list_stations",
                            location = c(6.29, 51.40),
                            time_interval = c(1990, 2020))
station_list <- slice(station_list, 1:25)

head(station_list)
```

2. Download weather data for the most promissing station on the list.
```{r}
weather_WES <- handle_gsod(action = "download_weather",
                          location = station_list$chillR_code[1],
                          time_interval = c(1990, 2020))
```


3. Convert the weather data into chillR format and save the data on your PC.
```{r}
cleaned_weather_WES <- handle_gsod(weather_WES)
cleaned_weather_WES[[1]][1000:1050,]


write.csv(station_list, "D:/! Ricardas Daten/Studium/Master/03. Semester (WiSe 24-25)/Tree phenology analysis in R/Logbook chapters/station_list_Wesel.csv", row.names = FALSE)
write.csv(weather_WES[[1]], "D:/! Ricardas Daten/Studium/Master/03. Semester (WiSe 24-25)/Tree phenology analysis in R/Logbook chapters/Wesel_raw_weather.csv", row.names = FALSE)
write.csv(cleaned_weather_WES[[1]], "D:/! Ricardas Daten/Studium/Master/03. Semester (WiSe 24-25)/Tree phenology analysis in R/Logbook chapters/Wesel_chillR_weather.csv", row.names = FALSE)
```

