---
title: "treeR_ch11"
author: "Ricarda Scholte-Reh"
date: "2024-10-12"
homepage: "https://Ricarda-Scholte-Reh.github.io"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
---

# Filling gaps in temperature records

## Exercises

1. Use chillR functions to find out how many gaps you have in this dataset (even if you have none, please still follow all further steps).
```{r}
rm(list=ls()) # clears the environment from unwanted values etc.
library(chillR)
library(tidyverse)
library(ggplot2)
library(tibble)
library(magrittr)

cleaned_weather_WES <- read.csv("D:/! Ricardas Daten/Studium/Master/03. Semester (WiSe 24-25)/Tree phenology analysis in R/Logbook chapters/Wesel_chillR_weather.csv")

#cleaned_weather_WES <- cleaned_weather_WES %>% make_all_day_table()

Wesel_QC <- fix_weather(cleaned_weather_WES)$QC

table(is.na(cleaned_weather_WES$Tmin))
Wesel_QC
```

2. Create a list of the 25 closest weather stations using the handle_gsod function.
```{r}
station_list <- handle_gsod(action = "list_stations",
                            location = c(6.29, 51.40),
                            time_interval = c(1990, 2020))
station_list <- slice(station_list, 1:25)
station_list
```

3. Identify suitable weather stations for patching gaps.
The stations 3, 5, 11 are suitable.

4. Download weather data for promising stations, convert them to ChillR format and compile them in a list
```{r}
patch_weather <- handle_gsod(action = "download_weather",
                             location = as.character(station_list$chillR_code[c(3, 5, 11)]),
                             time_interval = c(1990,2020)) %>% handle_gsod()
```

5. Use the patch_daily_temperatures function to fill gaps
```{r}
patched <- patch_daily_temperatures(weather = cleaned_weather_WES,
                                    patch_weather = patch_weather,
                                    max_mean_bias = 1,
                                    max_stdev_bias = 2)
```

6. Investigate the results - have all gaps been filled?
```{r}
patched$statistics[[1]]
patched$statistics[[2]]
patched$statistics[[3]]

#post_patch_stats <- fix_weather(patched)$QC

```

7. If necessary, repeat until you have a dataset you can work with in further analyses.
```{r}
station_list

patch_weather2 <- handle_gsod(action = "download_weather",
                             location = as.character(station_list$chillR_code[c(3, 4, 5, 7, 9, 11, 12, 13, 15, 16)]),
                             time_interval = c(1990,2020)) %>%
  handle_gsod()
```

```{r}
patched_2 <- patch_daily_temperatures(weather = cleaned_weather_WES,
                                    patch_weather = patch_weather2,
                                    max_mean_bias = 1,
                                    max_stdev_bias = 2)

patched_2$statistics[[1]]
patched_2$statistics[[6]]
patched_2$statistics[[7]]
```

```{r}
post_patch_stats_2 <- fix_weather(patched_2)$QC
post_patch_stats_2

patched_complete <- patched_2$weather %>% make_all_day_table()

Tmin_int <- interpolate_gaps(patched_complete[,"Tmin"])

patched_complete <- patched_complete %>% mutate(Tmin = Tmin_int$interp,
                              Tmin_interpolated = Tmin_int$missing)

Tmax_int <- interpolate_gaps(patched_complete[,"Tmax"])

patched_complete <- patched_complete %>% mutate(Tmax = Tmax_int$interp,
                              Tmax_interpolated = Tmax_int$missing)

fix_weather(patched_complete)$QC

fix_QC <- fix_weather(patched_complete)$QC
fix_QC
table(is.na(patched_complete$Tmin))

write.csv(patched_complete, "D:/! Ricardas Daten/Studium/Master/03. Semester (WiSe 24-25)/Tree phenology analysis in R/Logbook chapters/patched_weather_3_Wesel.csv", row.names = FALSE)
```
