---
title: "treeR_ch8"
author: "Ricarda Scholte-Reh"
date: "2024-10-12"
homepage: "https://Ricarda-Scholte-Reh.github.io"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
---

# Chapter 08 Making hourly temperatures

1.  Choose a location of interest, find out its latitude and produce plots of daily sunrise, sunset and day length.
    -   Location of interest: Dairy farm with own fodder cultivation (belongs to my uncle)
    -   Bislich, 46487 Wesel
    -   Latitude: 51°40'58.4"N Longitude: 6°29'46.4"E

```{r, echo = FALSE, results = 'hide', message = FALSE, warning = FALSE}
require(chillR)
require(ggplot2)
require(tidyr)
```

```{r}
Days_l8 <- daylength(latitude = 51.4, JDay = 1:365)
Days_df_l8 <- data.frame(JDay = 1:365,
                         Sunrise = Days_l8$Sunrise,
                         Sunset = Days_l8$Sunset,
                         Daylength = Days_l8$Daylength)

Days_df_l8 <- pivot_longer(Days_df_l8, cols=c(Sunrise:Daylength))

ggplot(Days_df_l8, aes(JDay, value)) +
        geom_line(lwd = 1.5) +
        facet_grid(cols = vars(name)) +
        ylab("Time of Day / Daylength [hours]") +
        xlab("Day of the year [JDay]") +
        theme_bw(base_size = 20)
```

2.  Produce an hourly dataset, based on idealized daily curves, for the KA_weather dataset (included in ChillR).

```{r}
hourtemp_KA <- stack_hourly_temps(KA_weather, latitude=50.4)
```

3.  Produce empirical temperature curve parameters for the Winters_hours_gaps dataset, and use them to predict hourly values from daily temperatures (this is very similar to the example above, but please make sure you understand what's going on).

```{r}
empi_curve_l8 <- Empirical_daily_temperature_curve(Winters_hours_gaps)

ggplot(data = empi_curve_l8[1:96, ], aes(Hour, Prediction_coefficient)) +
  geom_line(lwd = 1.3, col = "red") +
  facet_grid(rows = vars(Month)) +
  xlab("Hour of the day") +
  ylab("Prediction coefficient") +
  theme_bw(base_size = 20)
```

```{r}
coeffs_l8 <- Empirical_daily_temperature_curve(Winters_hours_gaps)
Winters_daily_l8 <- make_all_day_table(Winters_hours_gaps, input_timestep = "hour")
Winter_hours_l8 <- Empirical_hourly_temperatures(Winters_daily_l8, coeffs_l8)


# simplify the data for easier use
Winter_hours_l8 <- Winter_hours_l8[ , c("Year", "Month", "Day", "Hour", "Temp")]
colnames(Winter_hours_l8)[ncol(Winter_hours_l8)] <- "Temp_empirical"
Winters_ideal_l8 <- stack_hourly_temps(Winters_daily_l8, latitude = 38.5)$hourtemps
Winters_ideal_l8 <- Winters_ideal_l8[ , c("Year", "Month", "Day", "Hour", "Temp")]
colnames(Winters_ideal_l8)[ncol(Winters_ideal_l8)] <- "Temp_ideal"


# create the "triangular" data set to compare in the end
Winters_triangle_l8 <- Winters_daily_l8
Winters_triangle_l8[ , "Hour"] <- 0
Winters_triangle_l8$Hour[nrow(Winters_triangle_l8)] <- 23
Winters_triangle_l8[ , "Temp"] <- 0
Winters_triangle_l8 <- make_all_day_table(Winters_triangle_l8, timestep = "hour")
colnames(Winters_triangle_l8)[ncol(Winters_triangle_l8)] <- "Temp_triangular"


# the following loop will fill in the daily Tmin and Tmax values for every hour
for (i in 2:nrow(Winters_triangle_l8))
{if (is.na(Winters_triangle_l8$Tmin[i]))
  Winters_triangle_l8$Tmin[i] <- Winters_triangle_l8$Tmin[i - 1]
 if (is.na(Winters_triangle_l8$Tmax[i]))
   Winters_triangle_l8$Tmax[i] <- Winters_triangle_l8$Tmax[i - 1]}

Winters_triangle_l8$Temp_triangular <- NA


# Tmin is fixated to the 6th hour of each day
Winters_triangle_l8$Temp_triangular[which(Winters_triangle_l8$Hour == 6)] <-
  Winters_triangle_l8$Tmin[which(Winters_triangle_l8$Hour == 6)]


# Tmax is fixated to the 18th hour of each day
Winters_triangle_l8$Temp_triangular[which(Winters_triangle_l8$Hour == 18)] <-
  Winters_triangle_l8$Tmax[which(Winters_triangle_l8$Hour == 18)]


# the following loop will fill in the gaps between the min and max data in a linear way
Winters_triangle_l8$Temp_triangular <- interpolate_gaps(Winters_triangle_l8$Temp_triangular)$interp
Winters_triangle_l8 <- Winters_triangle_l8[ , c("Year", "Month", "Day", "Hour", "Temp_triangular")]


# merge all created data frames for easy display
Winters_temps_l8 <- merge(Winters_hours_gaps, Winter_hours_l8,
                          by = c("Year", "Month", "Day", "Hour"))
Winters_temps_l8 <- merge(Winters_temps_l8, Winters_triangle_l8,
                          by = c("Year", "Month", "Day", "Hour"))
Winters_temps_l8 <- merge(Winters_temps_l8, Winters_ideal_l8,
                          by = c("Year", "Month", "Day", "Hour"))


# convert the date into R's date format and reorganize the data frame
Winters_temps_l8[, "DATE"] <-
  ISOdate(Winters_temps_l8$Year,
          Winters_temps_l8$Month,
          Winters_temps_l8$Day,
          Winters_temps_l8$Hour)


Winters_temps_to_plot_l8 <-
  Winters_temps_l8[, c("DATE",
                    "Temp",
                    "Temp_empirical",
                    "Temp_triangular",
                    "Temp_ideal")]
Winters_temps_to_plot_l8 <- Winters_temps_to_plot_l8[100:200, ]
Winters_temps_to_plot_l8 <- pivot_longer(Winters_temps_to_plot_l8, cols=Temp:Temp_ideal)
colnames(Winters_temps_to_plot_l8) <- c("DATE", "Method", "Temperature")


ggplot(data = Winters_temps_to_plot_l8, aes(DATE, Temperature, colour = Method)) +
  geom_line(lwd = 1.3) + ylab("Temperature (°C)") + xlab("Date")
```
