---
title: "treeR_ch9"
author: "Ricarda Scholte-Reh"
date: "2024-10-12"
homepage: "https://Ricarda-Scholte-Reh.github.io"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
---

# Some useful tools in R

## Exercises

1. Based on the Winters_hours_gaps data set, use magrittr pipes and functions of the tidyverse to accomplish the following:
    a. Convert the data set into a tibble
    b. Select only the top 10 rows of the data set
    c. Convert the tibble to a long format, with separate rows for Temp_gaps and Temp
    d. Use ggplot2 to plot Temp_gaps and temp as facets (point or line plot)
    e. Convert the data set back to the wide format
    f. Select only the following columns: Year, Month, Day and Temp
    g. Sort the data set by the Temp column, in descending order

```{r}
rm(list=ls()) # clears the environment from unwanted values etc.
library(chillR)
library(tidyverse)
library(ggplot2)
library(tibble)
library(magrittr)

# Convert the data set into a tibble
hourtemps <- Winters_hours_gaps
hourtemps %>% as_tibble() %>% summary() 

# Select only the top 10 rows of the data set
hourtemps_l9 <- as_tibble(hourtemps[1:10, ])

# Convert the tibble to a long format, with separate rows for Temp_gaps and Temp
hourtemps_long_l9 <- hourtemps_l9 %>% pivot_longer(cols = Temp:Temp_gaps)


# Use ggplot2 to plot Temp_gaps and temp as facets (point or line plot)
ggplot(hourtemps_l9, aes(Hour, Temp_gaps)) +
  geom_point()

# Convert the data set back to the wide format
hourtemps_wide_l9 <- hourtemps_long_l9 %>% pivot_wider(names_from = name)

# Select only the following columns: Year, Month, Day and Temp
hourtemps_wide_l9 %>% select(c(Month, Day, Temp))

# Sort the data set by the Temp column, in descending order
hourtemps_wide_l9 %>% arrange(desc(Temp))
    
```

2. For the Winter_hours_gaps data set, write a for loop to convert all temperatures (Temp column) to degrees Fahrenheit.

```{r}
HT_l9.2 <- Winters_hours_gaps
HT_l9.2$Temp_F <- NA
for (i in 1:nrow(HT_l9.2))
  {HT_l9.2$Temp_F[i] <- (HT_l9.2$Temp[i] * 9/5) + 32}
```

3. Execute the same operation with a function from the apply family.

```{r}
HT_l9.3 <- hourtemps
func_Far_l9.3 <- function(x)
  {(x * 9/5) + 32}

for (i in 1:nrow(hourtemps))
  {HT_l9.3$Temp_F[i] <- sapply(HT_l9.3$Temp[i], func_Far_l9.3)}
```

4. Now use the tidyverse function mutate to achieve the same outcome

```{r}
HT_l9.2 <- hourtemps %>% mutate(Temp_F = (Temp * 9/5) + 32)
```

